---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r loadPackages, echo=FALSE, results='hide', message=FALSE}
# Load all required packages installing any that are missing.
requiredPackages <- c("dplyr", "ggplot2")
missingPackages <- !is.element(requiredPackages, installed.packages()[,1])
missingPackages <- requiredPackages[missingPackages]
if (length(missingPackages) != 0) install.packages(missingPackages)
lapply(requiredPackages, library, character.only = TRUE)
rm(requiredPackages, missingPackages)
```

## Loading and preprocessing the data
```{r loadData, echo=FALSE}
# Time when copy of activity.zip in this fork was downloaded.
dateDownloaded <- "Sat Jan 23 22:13:46 2016"

# Download data if not present and update time when the copy was downloaded.
if (!file.exists("activity.zip") && !file.exists("activity.csv")) {
    
    url = paste("https://d396qusza40orc.cloudfront.net/",
                "repdata%2Fdata%2Factivity.zip", sep = "")
    
    # Determine download.file method based on OS.
    if (.Platform$OS.type == "unix") {
        osMethod = "libcurl"
    } else {
        osMethod = "wininet"
    }
    
    download.file(url, "activity.zip", method = osMethod)
    dateDownloaded <- date()
    rm(url, osMethod)
}

# If .csv data file is not present if not unzip activity.zip
if (!file.exists("activity.csv")) {
    unzip("activity.zip")
}
```

File "activity.zip" was downloaded (on: **`r dateDownloaded`**) and unzipped to
"activity.csv".

```{r assignData}
activity <- tbl_df(read.csv("activity.csv"))
```



## What is mean total number of steps taken per day?

Any NAs in the data were omitted, the total steps taken per day were
calculated and plotted as a histogram:

```{r dailyActivity}
dailyActivity <- na.omit(activity) %>% group_by(date) %>%
    summarize(totalSteps = sum(steps))
```

```{r histogramSteps, echo=FALSE}
# Make a histogram of the total number of steps taken each day.
# Determine optimal bin number and corresponding breaks for histogram using
# the Freedman-Diaconis rule.
binNumber <- nclass.FD(dailyActivity$totalSteps)
binBreaks <- seq(min(dailyActivity$totalSteps),
                 max(dailyActivity$totalSteps),
                 length.out = binNumber + 1)

# Construct histogram
ggplot(dailyActivity) +
    geom_histogram(aes(totalSteps, ..count../sum(..count..) * 100),
                   origin = min(dailyActivity$totalSteps),
                   breaks = binBreaks,
                   col = "black",
                   fill= "grey") +
    scale_x_continuous(breaks = round(binBreaks)) +
    ggtitle("Normalized Histogram of Total Steps Per Day") +
    xlab("total steps per day") +
    ylab("percent of days [%]")
```

```{r meanMedianCalc, echo=FALSE}
# Calculate mean and median of the total steps taken per day
meanSteps <- mean(dailyActivity$totalSteps)
medianSteps <- median(dailyActivity$totalSteps)
```

The total number of steps taken per day have a mean of 
**`r format(meanSteps, digits = 7)`** and a median of **`r medianSteps`**.



## What is the average daily activity pattern?

To calculate the daily activity pattern, NAs were ommited, and the average
number of steps per interval was calculated and plotted as a line graph:

```{r intervalActivity}
intervalActivity <- na.omit(activity) %>% group_by(interval) %>%
    summarize(meanSteps = mean(steps))
```

```{r lineGraphActivity, echo=FALSE}
# Construct line graph with intervalActivity$interval as factor
labels <- intervalActivity$interval
fullHours <- labels[labels %% 100 == 0] 
ggplot(intervalActivity, aes(x = as.factor(interval), y = meanSteps, group=1)) +
    geom_line() +
    scale_x_discrete(breaks = fullHours) +
    ggtitle("Mean Steps Per 5-minute Interval Across All Days") +
    xlab("time interval (discrete)") +
    ylab("mean of steps") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r mostActiveInterval, echo=FALSE}
# Filter for maximum meanSteps and select corresponding interval
mostActiveInterval <- filter(intervalActivity,
                             meanSteps == max(intervalActivity$meanSteps))
mostActiveInterval <- mostActiveInterval$interval
```

The 5-minute interval **`r mostActiveInterval`** contains the maximum number
of steps on average across all the days.



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
